Draft 2

### 2020-07-30 Paper_V2.Rmd
ACS paper outlook

# Risk of infection from touching surfaces contaminated with SARS-CoV-2
## Abstract
## Introduction
SARS-CoV-2, the virus responsible for the COVID-19 pandemic, is transmitted via both direct (person-to-person) and indirect (via a contaminated environment) routes [@WHO2020;@CDC2020]. Direct transmission appears to be the leading route and occurs via prolonged exposures to respiratory droplets produced while talking, coughing, and sneezing. Infection control recommendations, based on the assumption that direct contact transmission is the leading route, include maintaining social/physical distances, wearing masks, case isolation, contact tracing, and quarantine [@Ferretti2020].

Despite knowledge that SARS-CoV-2 transmission is primarily through direct transmission, indirect transmission - or transmission occurring due to interaction with a contaminated environment – remains possible.   People infected with SARS-CoV-2 shed the virus into the environment, as evidenced by extensive environmental contamination detected on surfaces in cruise ships, hospitals, and XXX [@Ong2020; @Ye2020].  Infective coronavirus, including SARS-CoV-2, persists in the environment, with experimental evidence of persistence on surfaces for prolonged periods up to 9 days [@Kampf2020].  Viruses readily transfer from contaminated surfaces to the skin upon contact [@Julian2010; @Lopez2013; @Bidawid2004; @Rusin2002; @Ansari1991] and from the skin to the mouth and saliva of individuals [@Pitol2017; @Rusin2002].  Taken together, this evidence suggests surface contamination poses a risk for indirect SARS-CoV-2 transmission, similar to other respiratory viruses [@Boone2007].

Indirect transmission events of SARS-CoV-2, even if not the dominant transmission route, complicate infection control efforts using contact tracing.  Contact tracing efforts focus on identifying people sharing spaces coincidentally with cases [@Ferretti2020].  Indirect transmission could occur in the absence of shared spaces.  As an example, transmission of SARS-CoV-1 in a Taiwanese hospital led to nosocomial infection of 31 patients. In 6 (20%) of the patients, contact tracing failed to detect direct contacts with other SARS patients, although RNA of SARS-CoV-1 was detected throughout the hospital on drinking water buttons, beds, and chairs [@Chen2004].  Based on these findings, the authors suggest indirect transmission as the cause of transmission amongst a subset of those infected. [@Chen2004]. If indirect transmission is an important route for SARS-CoV-2, control strategies will necessarily have to integrate hand hygiene interventions and surface disinfection alongside contact tracing [@Mcclellan2020].

@@

SARS-CoV-2 QMRA model design to investigate the relative contribution of different transmission routes (contact, droplet, inhalation) for healthcare workers suggested that, although fomite-mediated transmission is not the main route of expossure, it accounts for 8% of the probability of infection without the use of PPE [@2020Jones].

@@

>>>>

# Risk of infection from touching surfaces contaminated with SARS-CoV-2
## Abstract

Objective. Use a stochastic mechanistic model to quantify the effect of surface [and hand] disinfection on SARS-CoV-2 infection risks. The results from this model can inform cleaning practices.

## Introduction
SARS-CoV-2, the virus responsible for the COVID-19 pandemic, is transmitted via both direct (person-to-person) and indirect (via a contaminated environment) routes [@WHO2020;@CDC2020]. Direct transmission appears to be the leading route and occurs via prolonged exposures to respiratory droplets produced while talking, coughing, and sneezing. Infection control recommendations, based on the assumption that direct contact transmission is the leading route, include maintaining social/physical distances, wearing masks, case isolation, contact tracing, and quarantine [@Ferretti2020].

Despite knowledge that SARS-CoV-2 transmission is primarily through direct transmission, indirect transmission - or transmission occurring due to interaction with a contaminated environment – remains possible.   People infected with SARS-CoV-2 shed the virus into the environment, as evidenced by extensive environmental contamination detected on surfaces in cruise ships, hospitals, and public spaces in urban areas such as bus stations and public squares [@Ong2020; @Ye2020; Abrahao2020].  Infective coronavirus persists in the environment, with experimental evidence of persistence on surfaces for prolonged periods up to 9 days [@Kampf2020].  Viruses readily transfer from contaminated surfaces to the skin upon contact [@Julian2010; @Lopez2013; @Bidawid2004; @Rusin2002; @Ansari1991] and from the skin to the mouth and saliva of individuals [@Pitol2017; @Rusin2002].  Taken together, this evidence suggests surface contamination poses a risk for indirect SARS-CoV-2 transmission, similar to other respiratory viruses [@Boone2007].

Indirect transmission events of SARS-CoV-2, even if not the dominant transmission route, complicate infection control efforts using contact tracing.  Contact tracing efforts focus on identifying people sharing spaces coincidentally with cases [@Ferretti2020].  Indirect transmission could occur in the absence of shared spaces.  As an example, transmission of SARS-CoV-1 in a Taiwanese hospital led to nosocomial infection of 31 patients. In 6 (20%) of the patients, contact tracing failed to detect direct contacts with other SARS patients, although RNA of SARS-CoV-1 was detected throughout the hospital on drinking water buttons, beds, and chairs [@Chen2004].  Based on these findings, the authors suggest indirect transmission as the cause of transmission amongst a subset of those infected [@Chen2004]. If indirect transmission is an important route for SARS-CoV-2, control strategies will necessarily have to integrate hand hygiene interventions and surface disinfection alongside contact tracing [@Mcclellan2020].

@@

SARS-CoV-2 QMRA model design to investigate the relative contribution of different transmission routes (contact, droplet, inhalation) for healthcare workers suggested that, although fomite-mediated transmission is not the main route of expossure, it accounts for 8% of the probability of infection without the use of PPE [@Jones2020].

@@

## Info for the Discussion
Frequency of hand to face contacts:
Quantified using video recordings 10 subjects wotking on an office during 3 hr. The subjects touched their eyes, nostrils, and lops an average of 16(11) times an hour @Nicas2008. Medical students touch their faces 23 times per hour, 44% of those contacts were with a mucous membrane @Kwok2015.
